OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-06-07 main.tex
Polished and reviewed the text. Two figure legends needs attention.

\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$,
François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb
1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Miguel
Beato\,$^{\text{\sfb 1,2}}$ and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}

@@

\begin{figure}
\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
\caption{A. Non linear relationship between the number of restriction
sites per bin and the total number of contacts per bin in T47D. B. Total
number of contacts per bin in chromosome 17 of T47D. Raw (brown) and
corrected (blue) signals are shown. The region that corresponds to the
first 20 Mbp of the chromosome is present in single copy in this clone of
T47D, explaining that the signal is approximately half compared to the
rest of the chromosome.  Also notice that the raw signal is noisier than
the corrected one.}
\label{fig:totals}
\end{figure}

@@

\begin{figure}[b]
	\centerline{\includegraphics[width=.45\textwidth]
{img/copy_number_figure2.pdf}}
	\caption{
Pearson correlation between total number of contacts per bin and
independent copy number estimation (COSMIC) for each of the methods
compared. Left panel T47D breast cancer cell line, right panel K562
leukemia cell line. The new proposal (in blue) outperforms the rest of
alternatives.}
\label{fig:copy_number}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=.50\textwidth]
  {img/correlation_aberrant_figure3.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotype. A and
B. Average changes compared to raw on the T47D and K562 sets. The bars
represent 95\% confidence intervals centered on the mean difference of the
correlation score between a given correction method and the raw data. The
brown dashed line indicates the value of the average score on raw matrices
(set to 0). C and D. ROC curves on the T47D and K562 sets. The areas under
the curve are indicated in the bottom right corner. The color code is the
same as panels A and B. The brown lines correspond to raw matrices. All
results in this figure are based on Pearson correlations between the
observed over expected counts.}
\label{fig:aberrant}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=.50\textwidth]
  {img/correlation_normal_figure4.pdf}}
\caption{
Results of the comparison between samples with normal karyotype.  A.
Average changes compared to raw on the mouse data set. The bars are as in
B. ROC curves on the mouse data set. The legend is as in
Figure~\ref{fig:aberrant}.}
\label{fig:normal}
\end{figure}

@@

Here we introduced \textit{OneD}, a fast computational method to normalize
Hi-C matrices. \textit{OneD} was developed ground up to address the need
to normalize data collected on samples with aberrant karyotypes, but it
applies seamlessly to the case of normal karyotypes. We showed that
\textit{OneD} performs significantly better than other methods when the
cells present karyotypic aberrations (Figure~\ref{fig:aberrant}), and that
it performs equally well as the best methods otherwise
(Figure~\ref{fig:normal}). We also showed that \textit{OneD} is
approximately as fast as \textit{ICE}, which makes it competitive from the
point of view of computational speed.

The originality of \textit{OneD} lies in that it projects all the biases
onto a single variable: the total amount of contacts per bin. This allows
greater running speed, while preserving a good performance on samples with
karyotypic aberrations. One of the reasons that \textit{OneD} is better
able to highlight the biological distinctions between samples is that it
does only corrects the copy number if specifically requested by the user.
Copy number variations treated as implicit biases have an
opaque effects on the

This raises the question whether variations of the copy number constitute
a biological signal or an artifact. Note that in the presence of
translocations, deletions and large scale rearrangements, correcting the
copy number is equally faulty as not correcting it. Indeed, in this case
the reference genome on which the matrix is drawn is incorrect and
changing the values of the pixels cannot account for this information.

Depending on the intention of the user, the effect of the copy number
should either be kept or removed. This is why \textit{OneD} does not perform
the correction by default, but allows the user to obtain a
diploid-equivalent Hi-C map computed from a hidden Markov model. The
resulting matrices have a near constant amount of contacts per bin, but
the artifacts caused by the misfit between the genome of the sample and
the reference genome are still present (for instance, the artifacts caused
by large scale inversions are not changed in any way).

Overall, \textit{OneD} constitutes a safe choice to normalize Hi-C matrices. If the
karyotype of the sample is aberrant, it enhances the biological variation.
If not, it performs at least equally well as other methods with an
advantage in terms of computational speed.

>>>>

\author[Vidal \textit{et~al}.]{Enrique Vidal\,$^{\text{\sfb 1,2}*}$,
François le Dily\,$^{\text{\sfb 1,2}}$, Javier Quilez\,$^{\text{\sfb
1,2}}$, Ralph Stadhouders\,$^{\text{\sfb 1,2}}$, Yasmina
Cuartero\,$^{\text{\sfb 1,2}}$, Thomas Graf\,$^{\text{\sfb 1,2}}$, Marc A.
Mart\'i Renom\,$^{\text{\sfb 1,2}}$, Miguel Beato\,$^{\text{\sfb 1,2}}$
and Guillaume J. Filion\,$^{\text{\sfb 1,2}}$}

@@

\begin{figure}
\centerline{\includegraphics[width=.45\textwidth]{img/figure1.pdf}}
\caption{Total amount of contacts per bin. A. Non linear relationship
between the number of restriction sites and the total number of
contacts per bin in T47D. B. Total number of contacts per bin on
chromosome 17 of T47D. Raw (brown) and corrected (blue) signals are shown.
The long arm of chromosome 17 (the region corresponding to 20-40 Mbp) is
present in four coipes, explaining that the signal is about twice higher
than for the short arm.}
\label{fig:totals}
\end{figure}

@@

\begin{figure}[b]
	\centerline{\includegraphics[width=.45\textwidth]
{img/copy_number_figure2.pdf}}
\caption{Pearson correlation between the total number of contacts per bin
and an independent estimation of the copy number (COSMIC). Left panel
T47D breast cancer cell line, right panel K562 leukemia cell line.}
\label{fig:copy_number}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=.50\textwidth]
  {img/correlation_aberrant_figure3.pdf}}
\caption{
Results of the comparison between samples with aberrant karyotypes. A and
B. Average changes compared to the raw data on the T47D and K562 sets. The
bars represent 95\% confidence intervals centered on the mean difference
of the correlation score between a given correction method and the raw
data. C and D. ROC curves on the T47D and K562 sets.  The areas under the
curve are indicated in the bottom right corner. The color code is the same
as panels A and B. The brown lines correspond to raw matrices. All results
in this figure are based on Pearson correlations between the observed over
expected counts.}
\label{fig:aberrant}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=.50\textwidth]
  {img/correlation_normal_figure4.pdf}}
\caption{
Results of the comparison between samples with normal karyotypes. A.
Average changes compared to the raw data on the mouse data set.  B. ROC
curves on the mouse data set. The legend is as in
Figure~\ref{fig:aberrant}.}
\label{fig:normal}
\end{figure}

@@

Here we introduced \textit{OneD}, a fast computational method to normalize
Hi-C matrices. \textit{OneD} was developed ground up to address the need
to normalize data from biological samples with aberrant karyotypes, but it
applies seamlessly to the case of normal karyotypes. We showed that
\textit{OneD} performs significantly better than other methods when the
cells present karyotypic aberrations (Figure~\ref{fig:aberrant}), and that
it performs equally well as the best methods otherwise
(Figure~\ref{fig:normal}). We also showed that \textit{OneD} is
approximately as fast as \textit{ICE}, which makes it competitive from the
point of view of computational speed.

The originality of \textit{OneD} lies in that it projects all the biases
onto a single variable: the total amount of contacts per bin. This allows
greater running speed, while preserving a good performance on samples with
karyotypic aberrations. One of the reasons that \textit{OneD} is better
able to highlight the biological distinctions between samples is that it
only corrects the copy number if specifically requested by the user. The
impact of copy number variations on normalization is rather opaque,
especially if they are treated as implicit biases
(Figure~\ref{fig:cnv_correction}). Treating them as explicit biases with
optional removal seems to be an overall safer strategy.

This raises the question whether variations of the copy number constitute
a biological signal or an artifact. If the biological sample contains
karytoypic aberrations, then its genome is grossly different the reference
so there can be no correct way to normalize the Hi-C signal. The proper
approach would be to use the actual genome of the biological sample as a
reference to construct the contact map. However, this strategy is
presently unfeasible because assembling mammalian genomes is still a hard
problem.

Depending on the intention of the user, the effect of the copy number
should either be kept or removed. This is why \textit{OneD} does not
perform the correction by default, but allows the user to obtain a
euploid-equivalent Hi-C map computed from a hidden Markov model. The
resulting matrices have a near constant amount of contacts per bin, but
the artifacts caused by the mismatch between the genome of the sample and
the reference genome are still present (for instance, the artifacts caused
by large scale inversions are not changed in any way).

Overall, \textit{OneD} constitutes a safe choice to normalize Hi-C
matrices. If the karyotype of the sample is aberrant, it enhances the
biological variation.  If not, it performs at least equally well as other
methods in terms of quality and of computational speed.
